---
title: "Test-script for applying a model to subsets of dataset"
author: "Andrew Nguyen"
date: "2016-March-07"
output: 
  pdf_document: 
    toc: true 
---
###Play script to show how to apply a stat model to a dataset

#load in libraries
```{r libraries}
#library(dplyr)
library(ggplot2)
library(plyr)
library(MASS)
```

#Loading in mock dataset
```{r load data}
babyURCH<-read.csv("20160307_playing_around.csv")
str(babyURCH)

#randomly generating expression data
babyURCH$gene_xp<-rnorm(length(babyURCH$Gene),100)+1000
hist(babyURCH$gene)

#visualize dataset
str(babyURCH)
#check replicates per treatment
ddply(babyURCH,.(Gene,Splice_variant,Day,acid),summarize,counts=length(rep))
```

#Fittign ANOVA model
```{r fitting ANOVA model}
#fitting models
#specifying the model with 3 way interaction
aovmod<-function(df){
  aov(gene_xp~Splice_variant*acid*Day,data=mutate(df))
}


#specifying the model with forward selection (can do backward or both too)
step.aovmod<-function(df){
  summary(stepAIC(aov(gene_xp~Splice_variant*acid*Day,data=mutate(df)),direction="forward"))
  #stepAIC(aov(gene_xp~Splice_variant*acid*Day,data=mutate(df)),direction="forward")
}

#applying the model for every gene
#.(Gene) splits the data based on gene
#models<-dlply(babyURCH,.(Gene),step.aovmod) 

#alternative way of doing it with lapply(this is better)
models2<-sapply(split(babyURCH,list(babyURCH$Gene)),step.aovmod)
models2
```

#Parsing output of model
```{r data parsing output}
#putting the model in a dataframe
full<-ldply(models2,data.frame)

#this line of code took the longest...grabbing the predictor names
#and just adding it to the dataframe!!
full$variables<-c(t(ldply(models2,row.names))[-1,])

#adding a new dataframe to indicate whether predictor 
#was significant at 0.05 level. Can toggle to lower if you want
full$sig<-ifelse(full$Pr..F.<0.05,"sig","nonsig")

#let's nicely visualize the dataframe!!
knitr::kable(full)

#writing out data frame into csv file :-)!
write.csv(full,"Test_output_model_dataframe.csv")
```

#Let's visualize gene "a" expression

Grab my ggplot defaults:
http://adnguyen.github.io/blog/2016/02/12/ggplot_defaults

The gray background and grids are so annoying.
```{r ggplot fig}
default<-theme_bw()+theme(text=element_text(size=30),axis.text=element_text(size=30), legend.text=element_text(size=28), panel.grid.major=element_blank(), panel.grid.minor.x = element_blank(), panel.grid = element_blank(), legend.key = element_blank())
library(Rmisc)

#grab averages
tc<-summarySE(babyURCH,measurevar="gene_xp",groupvars=c("Gene","Day","acid","Splice_variant"))
#grouping splice variant and acid levels together for coloring purposes
tc$grouping<-paste(tc$acid,tc$Splice_variant)
#visualize it
knitr::kable(tc)
pd <- position_dodge(0.5) #offset parameter in plotting points
#only looking at the "a" gene
a.gene<-subset(tc,tc$Gene=="a")

#mock figure
ggplot(a.gene,aes(y=gene_xp,x=factor(Day),color=grouping,shape=acid))+default+geom_errorbar(aes(ymin=gene_xp-sd,ymax=gene_xp+sd),size=1.5,width=.75,position=pd)+
geom_line(position=pd,size=1.5,stat="identity")+geom_point(position=pd,size=8)
```
You'll probably want boxplots because it better represents non-normal expression data (I think John assumed his expression data to be a negative binomial distribution).

###let's try a boxplot
```{r}
a.gene2<-subset(babyURCH,babyURCH$Gene=="a")
a.gene2$grouping<-paste(a.gene2$acid,a.gene2$Splice_variant)
ggplot(a.gene2,aes(y=gene_xp,x=factor(Day),color=grouping,fill=grouping))+default+geom_boxplot()
```

