% Copyright 2013 Christophe-Marie Duquesne <chmd@chmd.fr>
% Copyright 2014 Mark Szepieniec <http://github.com/mszep>
% 
% ConText style for making a resume with pandoc. Inspired by moderncv.
% 
% This CSS document is delivered to you under the CC BY-SA 3.0 License.
% https://creativecommons.org/licenses/by-sa/3.0/deed.en_US

\startmode[*mkii]
  \enableregime[utf-8]  
  \setupcolors[state=start]
\stopmode

\setupcolor[hex]
\definecolor[titlegrey][h=757575]
\definecolor[sectioncolor][h=397249]
\definecolor[rulecolor][h=9cb770]

% Enable hyperlinks
\setupinteraction[state=start, color=sectioncolor]

\setuppapersize [A4][A4]
\setuplayout    [width=middle, height=middle,
                 backspace=20mm, cutspace=0mm,
                 topspace=10mm, bottomspace=20mm,
                 header=0mm, footer=0mm]

%\setuppagenumbering[location={footer,center}]

\setupbodyfont[11pt, helvetica]

\setupwhitespace[medium]

\setupblackrules[width=31mm, color=rulecolor]

\setuphead[chapter]      [style=\tfd]
\setuphead[section]      [style=\tfd\bf, color=titlegrey, align=middle]
\setuphead[subsection]   [style=\tfb\bf, color=sectioncolor, align=right,
                          before={\leavevmode\blackrule\hspace}]
\setuphead[subsubsection][style=\bf]

\setuphead[chapter, section, subsection, subsubsection][number=no]

%\setupdescriptions[width=10mm]

\definedescription
  [description]
  [headstyle=bold, style=normal,
   location=hanging, width=18mm, distance=14mm, margin=0cm]

\setupitemize[autointro, packed]    % prevent orphan list intro
\setupitemize[indentnext=no]

\setupfloat[figure][default={here,nonumber}]
\setupfloat[table][default={here,nonumber}]

\setuptables[textwidth=max, HL=none]

\setupthinrules[width=15em] % width of horizontal rules

\setupdelimitedtext
  [blockquote]
  [before={\setupalign[middle]},
   indentnext=no,
  ]


\starttext

\section[andrew-d.-nguyen]{Andrew D. Nguyen}

\startblockquote
109 Carrigan Drive Burlington, VT 05405
\stopblockquote

\thinrule

\subsubsection[email-anbe642gmail.com]{Email: anbe642@gmail.com}

\subsubsection[github-adnguyen]{Github:
\useURL[url1][https://github.com/adnguyen][][adnguyen]\from[url1]}

\subsubsection[website-adnguyen.github.io]{Website:
\useURL[url2][https://adnguyen.github.io/][][adnguyen.github.io]\from[url2]}

\thinrule

\subsection[research-interests]{Research Interests}

I am broadly interested in understanding the ecological, evolutionary,
and genomic determinants of resiliency or susceptibility to climate
change.

\subsection[education]{Education}

\startdescription{2012-2017 (expected)}
  {\bf PhD, Biology}; University of Vermont (Burlington), Department of
  Biology

  {\em Thesis title: Evolutionary innovations of ants to thermally
  stressful environments}

  Advisors: Sara Helms Cahan, Nicholas J. Gotelli\crlf
  Committee: Brent L. Lockwood, Jill Preston
\stopdescription

\startdescription{2004-2009}
  {\bf BSc, Biology}; Drexel University (Philadelphia)
\stopdescription

\subsection[awards-and-grants]{Awards and Grants}

\startdescription{2016}
  Suiter Prize Travel Award - \$1,000
\stopdescription

\subsection[publications]{Publications}

\startitemize[n][stopper=.]
\item
  Helms Cahan S, {\bf Nguyen AD}, Stanton-Geddes J, Penick CA,
  HernáizHernández Y, DeMarco BB, Gotelli NJ. 2017. Modulation of the
  heat shock response is associated with acclimation to novel
  temperatures but not adaptation to climatic variation in the ants
  Aphaenogaster picea and A. rudis. Comparative Biochemistry and
  Physiology Part A: Molecular & Integrative Physiology 204:113--120.
\item
  {\bf Nguyen AD}, Gotelli NJ, Cahan SH. 2016. The evolution of heat
  shock protein sequences, cis-regulatory elements, and expression
  profiles in the eusocial Hymenoptera. BMC Evolutionary Biology 16:15.
\item
  Stanton-Geddes J, {\bf Nguyen A}, Chick L, Vincent J, Vangala M, Dunn
  RR, Ellison AM, Sanders NJ, Gotelli NJ, Cahan SH. 2016. Thermal
  reactionomes reveal divergent responses to thermal extremes in warm
  and cool-climate ant species. BMC Genomics 17:171.
\stopitemize

{\bf Forthcoming}

\startitemize[packed]
\item
  {\bf Nguyen AD}, DeNovellis K, Resendez S, Pustilnik J, Gotelli NJ,
  Parker JD, Helms Cahan S. (2016). Effects of desiccation and
  starvation on thermal tolerance and the cellular stress response in
  forest ants. In review at the Journal of Comparative Physiology B.
\stopitemize

\subsection[research-experience]{Research Experience}

\startdescription{2014}
  {\bf Research Associate}; University of Vermont

  Determined infection frequency of Trypanosome parasites in Kissing
  bugs that cause Chagas disease
\stopdescription

\startdescription{2010-2011}
  {\bf Technician}; Reaction Biology Corp

  High through put screening of potentially therapeutic small compounds.
\stopdescription

\startdescription{2009-2010}
  {\bf Technician}; Morphotek Inc.

  Developed therapeutic antibodies against Acute Myeloid Leukemia (AML).
\stopdescription

\startdescription{2007-2008}
  {\bf Co-op intern}; GlaxoSmithKline

  Studied red blood cell differentiation from mouse embryonic stem
  cells.
\stopdescription

\startdescription{2006-2007}
  {\bf Co-op intern}; Morphotek Inc.

  Developed neutralizing antibodies against Staphylococcal enterotoxin B
  (SEB).
\stopdescription

\subsection[skills]{Skills}

{\bf Computing:}

\startitemize[packed]
\item
  {\em Unix} -- General command line, shell bash scripting, and remote
  computing
\item
  {\em R} -- Data analysis and visualization
\item
  {\em (R)Markdown} -- Integrative word processing and technical
  reporting
\item
  {\em Python} -- Written scripts to parse genomic data
\item
  {\em Github} -- Reproducible science through version control and
  online notebooks
\item
  {\em HTML} -- Website development
\item
  {\em Phylogenetics} - Maximum likelihood (RAxML) and Bayesian
  (MrBayes)
\item
  {\em Geneious} -- Sequence analysis
\item
  {\em Microsoft Office} -- Data preparation, organization, and word
  processing
\stopitemize

{\bf Laboratory:}

\startitemize[packed]
\item
  RNA,DNA, and protein isolation\crlf
\item
  PCR and qPCR\crlf
\item
  Gel electrophoresis: polyacrylamide and agarose\crlf
\item
  Western blotting; immunohistochemistry\crlf
\item
  Cell culture (primary and established lines)\crlf
\item
  Flow Cytometry
\stopitemize

\subsection[external-reviewer]{External Reviewer}

Molecular Ecology\crlf
Insect Science

\subsection[research-presentations]{Research Presentations}

2017 : {\bf {\em Northern range limits of common forest ants is
reflected in trade-offs between basal and induced cold tolerances}},
Society of Integrative and Comparative Biology Conference, New Orleans,
LA (talk)

2016 : {\bf {\em Implementing strategies to achieve reproducible
research}}, BioLunch, University of Vermont, Department of Biology,
Burlington Vt (talk)

2015 : {\bf {\em Temperature adaptations in common woodland ants}},
BioLunch, University of Vermont, Department of Biology, Burlington Vt
(talk)

2014 : {\bf {\em Surviving in a warming world: thermal adaptation in
ants}}, BioLunch, University of Vermont, Department of Biology,
Burlington, Vt (talk)

2013 : {\bf {\em Impact of environmental stress on thermal tolerance in
Aphaenogaster picea}} EcoLunch, University of Vermont Department of
Biology, Burlington, Vt (Talk)

: {\bf {\em Physiological response to climate change in Aphaenogaster
picea}} , Northeast Natural History- Ant Ecology session, Springfield,
MA (Talk)

2012 : {\bf {\em Heat shock proteins and thermal tolerance in
Aphaenogaster picea}}, EcoLunch, University of Vermont Department of
Biology, Burlington, Vt (Talk)

: {\bf {\em Sequence and Cis-regulatory Evolution of Heat Shock Protein
hsp83, in Social Hymenoptera}}, International Union for the Study of
Social Insects- North American Section Meeting (IUSSI-NAS), Greensboro,
NC (Poster)

\placetable{{\bf {\em Heat shock proteins and thermal tolerance in
Aphaenogaster rudis}}, Aphaenophest, Petersham, MA (talk)}
\starttable[|l|]
\HL
\NC Conferences Attended
\NC\AR
\HL
\NC 2017
\NC\AR
\NC : Society of Integrative and Comparative Biology, New Orleans, LA
\NC\AR
\HL
\stoptable

2016 : Evolution, Austin, Texas 2014 Evolution, Raleigh, North Carolina

2014 : Molecular Biology and Evolution, Old San Juan, Puerto Rico

2013 : Northeast Natural History - Ant Ecology session, Springfield,
Massachusetts

2012 : International Union for the Study of Social Insects - North
American Section Meeting (IUSSI-NAS), Greensboro, North Carolina

\subsection[organizational-membership]{Organizational Membership}

\startblockquote
Society for Integrative & Comparative Biology (SICB)\crlf
American Society of Naturalists (ASN)\crlf
Society of Molecular Biology and Evolution (sMBE)\crlf
International Union for the Study of Social Insects (IUSSI)\crlf
Ecological Society of America (ESA)
\stopblockquote

\subsection[teaching-experience]{Teaching Experience}

2017 : {\bf Ecological Genomics} - Teaching Assistant, University of
Vermont\crlf
2016 : Invited Lecture, {\bf Evolution}, University of Vermont,

2015 : Invited Lecture, {\bf Evolutionary Biology} for non majors,
University of Vermont

2014 : Invited Lecture, {\bf Evolution}, University of Vermont

2014 : {\bf Exploring Biology} - Teaching Assistant, University of
Vermont

2013 : {\bf Cell and Molecular Biology} - Teaching Assistant, University
of Vermont

2012-2014 : {\bf Ecology and Evolution} - Teaching Assistant, University
of Vermont

2012 : {\bf Principles of Biology} - Teaching Assistant, University of
Vermont

\subsection[mentoring]{Mentoring}

{\bf Undergraduate Researchers:}

{\bf Ariana Maleki} and {\bf John Matthew Fisher} worked on developing
microsatellites for population genetic work on common woodland ants.

{\bf Kerri Pinder}, {\bf Skyler Resendez}, and {\bf Jeremy Pustilnik}
worked on how previous exposures to starvation and desiccation impact
thermal tolerance and underlying stress responses (using heat shock
proteins as a proxy).

{\bf Teddy Herriman} and {\bf Austin Sherburne} worked on identifying
potential morphological innovations that may temper and/or mitigate the
effects of heat stress.

{\bf Curtis A. Provencher} worked on how experimental warming affects
the stress levels of common woodland ants.

{\bf Megan Brown} and {\bf Jordan Zitnay} identified trade-offs between
constitutive and induced cold tolerances at the northern range boundary
in common forest ants, likely constraining their northern expansion.

\subsection[outreach]{Outreach}

2015 : Helper, Software Carpentry, University of Vermont

\starttyping
Helped participants troubleshoot code (Unix command line, R, Github)    
\stoptyping

2012 : Graduate Mentor, Ant Camp, University of Vermont

\starttyping
Aided in communicating and demonstrating ant collecting to high school students
\stoptyping

\stoptext
