---
title: "Simulating 95% confidence intervals"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---


# Load Libraries
```{r, warning=FALSE,message=FALSE}
library(tidyverse)
```


# Simulating 95% frequentist confidence intervals 

A good explanation [here](https://thestatsgeek.com/2020/11/21/interpretation-of-frequentist-confidence-intervals-and-bayesian-credible-intervals/):

**A 95% confidence interval is constructed such that if the model assumptions are correct and if you were to hypothetically repeat the experiment or sampling many many times, 95% of the intervals constructed would contain the true value of the parameter.** 

My own words: **The 95% confidence interval is when the true parameter is contained within the interval 95% of the time from constructing the 95% confidence interval from repeated experiments under the assumption of a correct model.** 

Let's gain intuition by what this means:    

1. Simulate data and do it a bunch of times\
2. Then calculate 95% confidence interval with say a t-test 
3. Determine how many times the true parameter (which we set in step 1) is in between the confidence intervals   



```{r}
#1) simulate data 
sim<-10000
#dataset size
n<-100
# sampel data with mean 10, sd =1 
x<-rnorm(n,mean=10,sd=1)
#fit t.test ; grab lower and upper confidence interval
#as.vector(c(t.test(x)$conf.int,t.test(x)$estimate))


## now simulate across sim 

#for loop is prob best 
#prep dataset
#d<-tibble(lower=rep(0,sim),upper=rep(0,sim),mean=rep(0,sim))
d<-array(0,dim=c(sim,2))

for (i in 1:sim){
  x<-rnorm(n,mean=10,sd=1)
  d[i,]<-as.vector(c(t.test(x)$conf.int))
}


#head(d)
d<-data.frame(d)
names(d)<-c("lower","upper")
knitr::kable(head(d))

#count how many times the lower and upper confidence interval is below true value of 10
d<-d|>
  mutate(out=1*(lower<10 & upper>10))
mean(d$out)

#cases where confidence interval is does not include true parameter 
d|>
  filter(out==0)|>
  head()
```

Additional notes:\

Cementing interpretation: When you have a single 95% CI on a single sample, it doesn't mean, that the population mean belongs to this particular interval with a particular probability. If you were to repeat the experiment many many times and calculate this interval on each fo the samples, then 95% of the repeated samples would have the true population mean. 

# Session info 

```{r}
sessionInfo()
```

