---
title: "Demo on non-collapsibility"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

# Load Libraries

```{r}
library(tidyverse)
library(marginaleffects)
library(gtsummary)
```

# Demonstrating non-collapsibility of odds ratios

Non-collapsibility refers to the fact that the conditional and marginal OR do not match. The conditional and marginal OR estimate agree in identity link regression model, so it is collapsible.

Simulate data first: outcome, treatment, and one continuous covariate

```{r}
########----##
n <- 5000
# 1) Simulate covariates (examples)
x1 <- rnorm(n, mean = 0, sd = 1)       # continuous
treatment <- rbinom(n, size = 1, prob = 0.5)  # binary indicator, 
# 2) Specify true effects (log-odds scale). Include factor via contrasts.
#    For factors, use model.matrix or include as factor in a data.frame and formula.
dat <- data.frame(x1, treatment)

# Build design matrix with contrasts automatically
X <- model.matrix(~ x1+treatment, data = dat)  # includes intercept and two dummies for x4

# Coefficients (choose your truth)
beta <- c(
  "(Intercept)" = -2.0,   # baseline log-odds
  "x1"          = 0.8,
  "treatment"          = -0.7
)

# 3) Linear predictor and probabilities
linpred <- as.numeric(X %*% beta)
p <- 1 / (1 + exp(-linpred))  # inverse-logit

# 4) Simulate binary outcome
y <- rbinom(n, size = 1, prob = p)

dat|>
  cbind(y)|>
  count(treatment,y)

#(213/2317)-(380/2090)

# Inspect prevalence and fit back a model to check recovery
#mean(y)
mod1<-glm(y ~ x1 + treatment, data = dat, family = binomial())
mod1|>
  tbl_regression(exponentiate = TRUE)

mod2<-glm(y ~ treatment, data = dat, family = binomial())
mod2|>
  tbl_regression(exponentiate = TRUE)
```

The regression coefficients for x3 (binary treatment effect) differ. (The true OR is 0.5)

# Now let's use g-computation

```{r}
# multiple variable logistic regression
avg_comparisons(mod1,newdata=dat,variables="treatment")

# marginal logistic regression 
avg_comparisons(mod2,newdata=dat,variables="treatment")
```

# Let's simulate more datasets and measure the OR and probability difference (ATE)

```{r}

n <- 5000

simdatORG<-function(n){
# 1) Simulate covariates (examples)
x1 <- rnorm(n, mean = 0, sd = 1)       # continuous
treatment <- rbinom(n, size = 1, prob = 0.5)  # binary indicator, 
# 2) Specify true effects (log-odds scale). Include factor via contrasts.
#    For factors, use model.matrix or include as factor in a data.frame and formula.
dat <- data.frame(x1, treatment)

# Build design matrix with contrasts automatically
X <- model.matrix(~ x1+treatment, data = dat)  # includes intercept and two dummies for x4

# Coefficients (choose your truth)
beta <- c(
  "(Intercept)" = -2.0,   # baseline log-odds
  "x1"          = 0.8,
  "treatment"          = -0.7
)

# 3) Linear predictor and probabilities
linpred <- as.numeric(X %*% beta)
p <- 1 / (1 + exp(-linpred))  # inverse-logit

# 4) Simulate binary outcome
y <- rbinom(n, size = 1, prob = p)

# Inspect prevalence and fit back a model to check recovery
#mean(y)
mod1<-glm(y ~ x1 + treatment, data = dat, family = binomial())
#mod1$coefficients[3]
mod2<-glm(y ~ treatment, data = dat, family = binomial())
#mod2$coefficients[2]#
#g-comput
g1<-avg_comparisons(mod1,newdata=dat,variables="treatment")
#g1$estimate
g2<-avg_comparisons(mod2,newdata=dat,variables="treatment")
#g2$estimate


outtbl<-tibble(condbiasOR=exp(mod1$coefficients[3]),margbiasOR=exp(mod2$coefficients[2]),cmm=condbiasOR-margbiasOR,condbiasg=g1$estimate,margbiasg=g2$estimate,atecmm=condbiasg-margbiasg)
return(outtbl)
}

#simulation
sim0.1<-tibble(n=1:1000)|>
  group_by(n)|>
  do(out=simdatORG(n=1000))|>
  unnest(out)
#sim0.1

comptbl<-sim0.1|>
  select(-n)|>
  tbl_summary(label = list(condbiasOR~"Conditional OR",margbiasOR~"Marginal OR",condbiasg~"Conditional ATE",margbiasg~"Marginal ATE",cmm~"Condition-Marginal OR",atecmm~"Condition-Marginal ATE"))|>
  as_gt()
comptbl

#saving figures
#gtsave(comptbl,filename="noncollapsibility_of_OR_treatment_covariate_model_vs_marginal_model.png")
#gtsave(comptbl,filename="noncollapsibility_of_OR_treatment_covariate_model_vs_marginal_model.docx")
```

Across 1,000 simulations, the OR differ between a conditional OR vs marginal OR, but ATE remains the same between models (multivariable vs marginal model).

# Sessioninfo

```{r}
sessionInfo()
```
