---
title: "Bayesian clinical trial simulations"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

# Load library

```{r load libraries , warning=FALSE,message=FALSE}
library(magrittr)
library(tidyverse)
library(brms)
library(purrr)
library(marginaleffects)
library(patchwork)
library(cowplot)
library(gt)
library(gtsummary)
library(ggbeeswarm)
oh_cols<- c('#50BECB',	'#46A6B2',	'#65C9D5',	'#97D3DC',	'#CDEBF0',	'#EDE668',	'#AA1E2D',	'#E4E5E3',	'#F26828',	'#DC5C1D',	'#F89C70',	'#FDCEB0',	'#2A3C47',	'#18272F',	'#404C58',	'#646A74',	'#C3C3C8',	'#74308C')

brms_summary <- function(x) {
  posterior::summarise_draws(x, "mean", "sd",  ~quantile(.x, probs = c(.025,0.8,0.95)))
}

options(mc.cores = parallel::detectCores())

```

# Intro

This script showcases how to conduct power analyses for Bayesian clinical trial designs. I'll explore equivalence, non-inferiority, superiority/efficacy designs.

# Establishing Bayesian priors, mcid, and similarity interval

As an example, I'd like to show the prior distribution first.

-   the MCID is log(0.8) or `r log(1/1.5)`\
-   the similarity interval is \[`r log(1/1.5)` - `r log(1.5)`\]

```{r graphing bayesian priors}
#priors
#lets see range of OR in log scale
ordat<-round(tibble(or=c(0.1,.2,0.5,0.667,1,1.5,2,4,5,10),logor=log(or)),3) # +/- 1.61

prior.plot<-ggplot(data = tibble(x = -2.3:2.3), aes(x)) +annotate(xmin=log(1/1.5),xmax=log(1.5),ymin=0,ymax=0.5,'rect',alpha=.5)+
  stat_function(fun = dnorm, n = 101, args = list(0,log(5)/qnorm(0.975)),aes(colour="Skeptical"),linewidth=2) +theme_bw()+xlab("Log-odds ratio")+ylab("Density")+scale_x_continuous(limits=c(-2.3,2.3),labels =ordat$logor,breaks=ordat$logor,sec.axis = dup_axis(name="Odds ratio",labels=ordat$or))+
  scale_colour_manual(breaks=c("Skeptical"),values=c("#97D3DC"),name="Prior")+theme(legend.position="top")+annotate("text",x=0,y=.02,label="Similarity interval")
prior.plot


```

# Power analysis

Bayesian power = probability of reaching trial goals

## fitting initial model

This part of the script will be useful for the real analysis, which includes fitting the logistic model, checking the model, and testing hypotheses.

```{r power analysis}
n=100 # sample size of patients total 

#simulating the dataset
d<-tibble(treatment=rbinom(n,1,.5))|>
  mutate(y=ifelse(treatment ==0,rbinom(n,1,.19),
                            rbinom(n,1,.19)))
d|>
  group_by(treatment)|>
  count(y)

#fit a brms model with skeptical prior
#fit <-brm(data = d,
#      family = bernoulli(),
#      y~treatment,
#      prior = c(prior(normal(0,log(5)/1.96), class = #b)),chains=4,cores=4)
#saveRDS(fit,file="Power_analysis_simulations_outputs/Initial_fit_Bayesian_model.rds")
fit<-readRDS("Power_analysis_simulations_outputs/Initial_fit_Bayesian_model.rds")

plot(fit) # see chains
print(fit) # check rhat, effective sample size, and model output
prior_summary(fit)
###posterior predictive checks 
pp_check(fit,type="stat",stat="mean")
pp_check(fit,stat="mean",ndraws=50)

#get priors, if interested
#get_prior(fit)
#marginal effects

```

### Hypothesis testing of posterior distribution

```{r testing hypotheses}
#test hypotheses
# test treatment <0, overall benefit
ob<-hypothesis(fit,"treatment<0",alpha=0.1)
ob
###test for similarity ; within log(0.8)-log(1.2)
si1<-hypothesis(fit,"treatment < log(1.5)",alpha=0.05)
si2<-hypothesis(fit,"treatment <log(0.667)",alpha=0.05)
si1$hypothesis$Post.Prob-si2$hypothesis$Post.Prob

# dobule check my understanding by manually calculating probs
samples<-as.data.frame(fit)
n<-length(samples$b_treatment)

#manual for overall benefit
sum(samples$b_treatment<0)/n
#similarity interval 
sum(samples$b_treatment>log(0.667) & samples$b_treatment<log(1.5))/n

#### hypothesis function and manual calculations are in agreement #####

##plotting out the treatment effect 
subs<-tibble(treatment=samples$b_treatment)|>
  mutate(Prior="Skeptical")
mean.paste<-paste(round(exp(mean(subs$treatment)),3)," odds ratio\n",round(mean(subs$treatment),3),"  log odds ratio",sep="")
mp<-round(mean(subs$treatment),3)

ggplot(subs,aes(x=treatment,..scaled..))+geom_density(fill='#46A6B2',alpha=.5)+theme_bw()+xlab("Treatment effect (log odds ratio)")+
  scale_x_continuous(limits=c(-1.61,1.61),labels =c(-2.302,ordat$logor),breaks=c(-2.302585,ordat$logor),sec.axis = dup_axis(name="Treatment effect (odds ratio)",labels=c(0.1,ordat$or)))+
  scale_y_continuous(expand=c(0,0),limits=c(0,1.3))+
  annotate(xmin=log(1/1.5),xmax=log(1.5),ymin=0,ymax=1.3,'rect',alpha=.5)+
  annotate("text",x=-1,y=1.15,label = mean.paste)+
  annotate("text",x=log(1/1.5),y=1.2,label="Similarity interval",hjust=-.05)+ylab("Density")+
  geom_vline(xintercept = 0,linewidth=1,lty="dotdash")

```

## Simulation for power analysis: Similarity and Non-inferiority

The simulation takes a long time, so I ran and saved them.

```{r simulation and fit}
####fit simulate and fit
#n=50
seed<-1

simfit <- function(seed, n) {
  
  set.seed(seed)
  
  d<-tibble(treatment=rbinom(n,1,.5))|>
  mutate(y=ifelse(treatment ==0,rbinom(n,1,.19),
                            rbinom(n,1,.19)))
  outp<-update(fit,
         newdata = d,
         seed = seed,cores=4,recompile=FALSE,sample_prior="no",refresh=0) 
  ot<-outp|>
    brms_summary()|>
  data.frame()|>
  filter(variable=="b_treatment")
  
  si1<-hypothesis(outp,"treatment < log(1.5)",alpha=0.05)
  si2<-hypothesis(outp,"treatment <log(0.6666)",alpha=0.05)
  ot$similarity<-si1$hypothesis$Post.Prob-si2$hypothesis$Post.Prob
  ot$similarity_below15<-si1$hypothesis$Post.Prob
  
  #benefit<-hypothesis(outp,"treatment <log(1)",alpha=0.05)
  #ot$benefit<-benefit$hypothesis$Post.Prob
  #ot$harm<- (1-ot$benefit)
  return(ot)
}


#######################################################
n_sim<-100 # number of simulations
#simulation here, didn't run
#sim2 <-tibble(expand.grid(seed=1:n_sim,ss=c(100,200,400,600,800)))|>
#  group_by(seed,ss)|>
#  do(out=simfit(seed=.$seed,n=.$ss))|>
#  unnest(out)

#mutate(out = pmap(list(seed, ss), ~ simfit(seed = ..1, n = ..2))) %>%

#saveRDS(sim2,file="Power_analysis_simulations_outputs/Non-inferiority_andsimilarity_Bayesian_simulation_100trials_ss50-1000_log5over1.96prior_1.5OR.rds")

sim2<-readRDS(file="Power_analysis_simulations_outputs/Non-inferiority_andsimilarity_Bayesian_simulation_100trials_ss50-1000_log5over1.96prior_1.5OR.rds")
#calculating bayesian power for noninferiority and similarity
sim2<-sim2|>
  group_by(ss)|>
  mutate(prbelow15=sum(if_else(similarity_below15>0.95,1,0))/length(similarity_below15),facet=paste("Pr(Non-inferiority>0.95)=",prbelow15,sep=""),
         simprob=sum(if_else(similarity>.95,1,0)/length(similarity)),
         facetsim=paste("Pr(Similarity>0.95)=",simprob,sep=""))|>
  droplevels()
head(sim2)
sim2|>
  count(ss)



###############Non-inferiority trial

nfplot<-ggplot(sim2,aes(y=similarity_below15,x=factor(ss)))+geom_quasirandom(shape=21,fill='#50BECB',alpha=.5,size=2)+scale_y_continuous(breaks=seq(0,1,.1),labels=seq(0,1,.1))+geom_hline(yintercept = .95,lty="dotdash")+xlab("Sample size")+ylab("Non-inferiority: Pr(OR<1.5)")+facet_wrap(~facet,scale="free_x",nrow=1)+theme_bw()
nfplot

#ggsave(nfplot,filename="Power_analysis_simulations_outputs/Non-inferiority_simulation_ss50-1000-log5over1.96prior_1.5OR.png",width=10,height=5,units="in",dpi=600)


##similarity trial
simfig<-ggplot(sim2,aes(y=similarity,x=factor(ss)))+geom_quasirandom(shape=21,fill='#50BECB',alpha=.5,size=2)+scale_y_continuous(breaks=seq(0,1,.1),labels=seq(0,1,.1))+geom_hline(yintercept = .95,lty="dotdash")+xlab("Sample size")+ylab("Similarity: Pr(0.66< OR <1.5)")+facet_wrap(~facetsim)+theme_bw()
simfig

#ggsave(simfig,filename="Power_analysis_simulations_outputs/equivalence_simulation_ss50-1000-log5over1.96prior_1.5OR.png",width=8,height=4,units="in",dpi=600)

```

# Efficacy trial simulation

```{r}
#####simulation function for efficacy 
simfitefficacy <- function(seed, n) {
  
  set.seed(seed)
  
  d<-tibble(treatment=rbinom(n,1,.5))|>
  mutate(y=ifelse(treatment ==0,rbinom(n,1,.2),
                            rbinom(n,1,.1)))
  
    outp<-update(fit,
         newdata = d,
         seed = seed,cores=4) 
  ot<-outp|>
    brms_summary()|>
  data.frame()|>
  filter(variable=="b_treatment")
  
  benefit<-hypothesis(outp,"treatment <log(1)",alpha=0.05)
  ot$benefit<-benefit$hypothesis$Post.Prob
  ot$harm<- (1-ot$benefit)
  return(ot)
}

###simulation
#n_sim<-100 # number of simulations
# --simulation, didn't run
#sim3 <-tibble(expand.grid(seed=1:n_sim,ss=c(100,200,400,600,800)))|>
#  mutate(out = pmap(list(seed, ss), ~ simfitefficacy(seed = ..1, n = ..2))) |>
#  unnest(out)

#saveRDS(sim3,file="Power_analysis_simulations_outputs/EFFICACY_Bayesian_simulation_100trials_ss50-1000_log5over1.96prior_1.5OR.rds")
sim3<-readRDS("Power_analysis_simulations_outputs/EFFICACY_Bayesian_simulation_100trials_ss50-1000_log5over1.96prior_1.5OR.rds")

sim3<-sim3|>
  group_by(ss)|>
  mutate(probeff=sum(if_else(benefit>0.95,1,0))/length(benefit),
         facet=paste("Pr(Efficacy>0.95)=",probeff))



fig_eff<-ggplot(sim3,aes(x=factor(ss),y=benefit))+geom_quasirandom(size=3,shape=21,fill='#50BECB',alpha=.5)+scale_y_continuous(breaks=seq(0,1,.1),labels=seq(0,1,.1))+xlab("Sample size")+ylab("Efficacy: Pr(OR<1)")+geom_hline(yintercept = 0.95,lty="dotdash",linewidth=1)+theme_bw()+facet_wrap(~facet,nrow=1)+theme(axis.text.x = element_text(angle = 45, hjust = 1))
fig_eff
#ggsave(fig_eff,filename="Power_analysis_simulations_outputs/efficacy_simulation_ss50-1000-log5over1.96prior_1.5OR.png",width=8,height=4,units="in",dpi=700)


```

# Posterior predictive distributions -assess whether a trial is likely to succeed

1.  Define success criteria-\> Pr(OR \<0 )\>=0.95
2.  Fit a Bayesian model (say halfway through the trial)
3.  Posterior predictive distribution

$$p(\tilde{y}|data) = \int p(\tilde{y}|\theta) p(\theta |data) d\theta$$ 4. Compute predictive probabilty of success (PPOS = Pr(Success at end \| current data))

Here, simulate many future trial completions from posterior predictive distribution

5.  Decision rules: if PPOS is high,continue, if it is low, stop

```{r initial model fit, eval=FALSE }
#rbinom(n=1,size=1,prob=.4)
#workflow

#conditions
#efficacy trial
# trial with 100 total patients, target is 200 patients total ; 100 per treatment
# control = .2, treatment = .1
n=100
d<-tibble(treatment=rbinom(n,1,.5))|>
  mutate(y=ifelse(treatment ==0,rbinom(n,1,.2),
                            rbinom(n,1,.1)))

#write.csv(d,"Power_analysis_simulations_outputs/d_dataset_randomtreatment_ybinary.csv")
d<-read.csv("Power_analysis_simulations_outputs/d_dataset_randomtreatment_ybinary.csv")|>
  select(-X)

#fit model with 100 patients
fit <-brm(data = d,
      family = bernoulli(),
      y~treatment,
      prior = c(prior(normal(0,log(5)/1.96), class = b)),chains=4,cores=4)
#summary(fit)
#saveRDS(fit,"Power_analysis_simulations_outputs/00_simulated_dataset_original_brms_N100_logistic_regression.rds")

#fit<-readRDS("Power_analysis_simulations_outputs/00_simulated_dataset_original_brms_N100_logistic_regression.rds")
summary(fit)
#Hypothesis testing here: Is there a treatment effect? 
#fithyp<-hypothesis(fit,"treatment<log(1)")
#fithyp$hypothesis$Post.Prob

#get posterior probabilities and link them up with original dataset 
# N goes from 100 to 200 

newdat<-tibble(treatment=rbinom(100,1,.5))|>
  arrange(treatment)

pdraws<- posterior_predict(fit,summary=FALSE,newdata=newdat,ndraws=100)|>
  data.frame()##1st column is control, 2nd column is treatment 
names(pdraws)<-c("control","treatment")
numdat<-newdat|>
  count(treatment)
  #mutate(y=if_else(treatment==0,sample(pdraws$control,size=1)))
newdat2<-newdat|>
  mutate(y=c(pdraws[1:numdat[[1,2]],1],pdraws[1:numdat[[2,2]],1]))
fulld<-rbind(d,newdat2)

#refit model using update 100 times
upfit<-update(fit,
         newdata = fulld,cores=4) 
# determine which ones successful and calculate prob of success 
hyp1<-hypothesis(upfit,"treatment<log(1)")
hyp1$hypothesis$Post.Prob

#now i need to construct a function to repeat many many times 
```

## Code to simulate new patients based on the current model

```{r eval=FALSE}
############
#first create a datasets from already collected + posterior predicted outcomes
dat<-list()

n_iter <- 1000
start_time <- proc.time()[["elapsed"]]

for (i in seq_len(n_iter)) {
  iter_start <- proc.time()[["elapsed"]]

  # === your work ===
  ndat<-tibble(treatment=c(0,1))
  pdraws <- posterior_predict(fit, summary = FALSE,newdata = ndat, ndraws = 200) |>
    data.frame()            # 1st col: control, 2nd: treatment
  names(pdraws) <- c("control", "treatment")
  
  
  newdat<-tibble(treatment=rbinom(100,1,.5))|>
    arrange(treatment)
  numdat<-newdat|>
    count(treatment)
  newdat<-newdat|>
    mutate(y=c(pdraws[1:numdat[[1,2]],1],pdraws[1:numdat[[2,2]],1]))
  fulldsim<-rbind(d,newdat)
  
  dat[[i]] <- fulldsim

  # === progress print ===
  iter_elapsed <- proc.time()[["elapsed"]] - iter_start
  total_elapsed <- proc.time()[["elapsed"]] - start_time
  avg_per_iter <- total_elapsed / i
  eta <- avg_per_iter * (n_iter - i)

  cat(sprintf("Iter %d/%d | iter: %.2fs | total: %.2fs | ETA: %.2fs\n",
              i, n_iter, iter_elapsed, total_elapsed, eta))
}
```

## Simulation and PPOS

```{r simulation }
#head(dat)
#from the list of datasets, make it into a dataframe 
#fdat<-bind_rows(dat,.id="source_id") #ocmment out, not run

#write.csv(fdat,"Power_analysis_simulations_outputs/00_simulated_dataset_1000_brms_N100_logistic_regression_posteriorpredict_N50_rbind.csv")
fdat<-read.csv("Power_analysis_simulations_outputs/00_simulated_dataset_1000_brms_N100_logistic_regression_posteriorpredict_N50_rbind.csv")
head(fdat)

#now I can iterate on each partition of the dataset by source_id
#write a function to update

updatebrm<-function(data){
  data=data
  outp<-update(fit,
         newdata = data,
         cores=4) 
  ot<-outp|>
    brms_summary()|>
  data.frame()|>
  filter(variable=="b_treatment")
  ot$benefit<-hypothesis(outp,"treatment <log(1)",alpha=0.05)$hypothesis$Post.Prob
  return(ot)
}

#now lets iterate 
#simp1<-fdat[1:400,]|> # testing on 2 simulations

#the simulation code here, not run, due to time
#simp1<-fdat|>
#  group_by(source_id)|>
#  do(output=updatebrm(data=.))|>
#  unnest(output)
#simp1
#write.csv(simp1,row.names=FALSE,"Power_analysis_simulations_outputs/00_simulated_dataset_1000_brms_N100_logistic_regression_posteriorpredict_N50_rbind_output_dataset_benefit.csv")
simp1<-read.csv("Power_analysis_simulations_outputs/00_simulated_dataset_1000_brms_N100_logistic_regression_posteriorpredict_N50_rbind_output_dataset_benefit.csv")

ggplot(simp1,aes(y=benefit,x=variable))+geom_quasirandom()+xlab("Treatment")+ylab("Posterior predictive success (Pr(OR<1))")+scale_y_continuous(labels=seq(0,1,.1),breaks=seq(0,1,.1))+geom_hline(yintercept = .8,lty="dotdash")
#

sum(if_else(simp1$benefit>0.9,1,0))/length(simp1$benefit)



cutoff <- 0.9
d <- density(simp1$benefit)
dens_df <- tibble(x = d$x, y = d$y)

fig1.1<-ggplot(dens_df, aes(x, y)) +
  geom_line(color = '#18272F') +
  geom_ribbon(
    data = filter(dens_df, x > cutoff),
    aes(ymin = 0, ymax = y),
    fill = '#65C9D5', alpha = 0.5
  ) +
  xlab("Pr(OR<1)") + ylab("Density")+theme_bw()+scale_x_continuous(limits=c(.3,1),breaks=seq(0,1,.1),labels=seq(0,1,.1))
fig1.1
#ggsave(fig1.1,filename="Power_analysis_simulations_outputs/00_fig_posterior_Predictive_success_fig_1000simulations.png",width=8,height=5,dpi=900,units="in")
```

# session info

```{r session info}
sessionInfo()
```
