---
title: "Regression continuity designs"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

# Load libraries

```{r ,warning=FALSE,message=FALSE}
library(tidyverse)
library(brms)
library(marginaleffects)
```

# Introduction: Regression discontinuity

Reviewing regression discontinuity from sources:

-   https://www.bayes-pharma.org/wp-content/uploads/2019/06/Session-5.4-Sara-Geneletti.pdf

The key to RD is the threshold acts like a randomizing device or an instrumental variable. Units close to the threshold are assumed to satisfy conditional exchangeability.

Assumptions:

1.  Unconfoundedness
2.  indpendence of guidelines
3.  monotonicity
4.  Continuity

1-3 are instrumental variable assumptions

Sharp RD estimator:

$$ATE = E(Y|Z=1) - E(Y|Z=0) - E(T|Z)=0 = \Delta_\beta = \beta_{0a}-\beta_{0b}$$ Estimand = sharp jump in regression function at the cutoff.

$$Y_i = \alpha + \tau D_i + \beta_1 (X_i - c) + \beta_2 D_i (X_i - c) + \epsilon_i$$

-   $\tau$ is the estimator - treatment effect of discontinuity
-   $D_i = 1(X_i \geq  c)$\
-   

# Single group regression discontinuity

Simulate data under 2 scenarios and then fit bayesian regression models:

-   No treatment effect
-   Treatment effect

## No treatment effect

```{r singless}
#simulate data 
n<-5000
x<-runif(n,-10,10)
treatment<-if_else(x<0,0,1)
treatment_fac<-if_else(x<0,"before","after")
cutoff<-0


beta0<-50 # intercept
beta1<-1.25 #slope at cutoff
tau<-0 #jump at cutoff
beta2<- 0  #slope change above cutoff

y<-beta0+beta1*(x-cutoff)+tau*treatment+beta2*treatment*(x-cutoff)+rnorm(n=n,mean=0,sd=2)

dat<-tibble(x=x,y=y,treatment=treatment_fac)|>
  mutate(x_c=x-cutoff,D=if_else(x>=cutoff,1,0))

ggplot(dat,aes(x=x,y=y,colour=factor(treatment)))+geom_point(alpha=.2)+stat_smooth(method="lm")

#let's fit a model now to see if we recover the estimate 

#set up prior = N(0,5) for fixed effects
# set up intercept prior as N(0,30)
priorset<-c(prior(normal(0,30),class="Intercept"),
            prior(normal(0,10),class=b))

#fit model 
#mod1<-brm(y~D + x_c + D:x_c,prior = priorset,data=dat,family = gaussian())
#saveRDS(mod1,"24_regression_discont_brms_model_withNOtreatmenteffect.rds")
mod1<-readRDS("24_regression_discont_brms_model_withNOtreatmenteffect.rds")
prior_summary(mod1) # checking priors in the model 
summary(mod1) # getting results or model output, examine effective sample size and rhat
#check chains 
plot(mod1) # chains well mixed

pp_check(mod1, type = "dens_overlay",ndraws=50)
#pp_check(mod1, type = "scatter",ndraws=50)# good for RD 
#pp_check(mod1, type = "intervals",ndraws=50)   # expected y vs observed y
pp_check(mod1, type = "ecdf_overlay",ndraws=50)
pp_check(mod1, type = "stat_2d",ndraws=50)     # good for RD with x vs y

```

## With a treatment effect

```{r with treatment effect}
tau2<-2.5 #jump at cutoff

y2<-beta0+beta1*(x-cutoff)+tau2*treatment+beta2*treatment*(x-cutoff)+rnorm(n=n,mean=0,sd=2)
dat2<-tibble(x=x,y=y2,treatment=treatment_fac)|>
  mutate(x_c=x-cutoff,D=if_else(x>=cutoff,1,0))

ggplot(dat2,aes(x=x,y=y,colour=factor(treatment)))+geom_point(alpha=.2)+stat_smooth(method="lm")

# fit model ; use same priors
#mod2<-brm(y~D + x_c + D:x_c,prior = priorset,data=dat2,family = gaussian())
#write_rds(mod2,file="24_regression_discont_brms_model_withtreatmenteffect.rds")
mod2<-readRDS("24_regression_discont_brms_model_withtreatmenteffect.rds")
#prior_summary(mod2) # checking priors in the model 
summary(mod2) # getting results or model output, examine effective sample size and rhat
#the D parameter matches what we set above, 2.5


#plot with conditional effects
#plot(conditional_effects(mod2,"x_c",conditions=list(D=c(0,1))))
#not really what i want
```

### Different way to plot

```{r different way to plot}
#make new dataset

newdat <- expand.grid(
  x_c = seq(min(dat$x_c), max(dat$x_c), length.out = 200),
  D = c(0, 1)
)
#get estimates
newdat$y_hat <- fitted(mod2, newdata = newdat)[, "Estimate"]


ggplot() +
  geom_point(data = dat2, aes(x = x_c, y = y, color = factor(D)), alpha = 0.4) +
  geom_line(data = newdat, aes(x = x_c, y = y_hat, color = factor(D)), size = 1.2) +
  scale_color_manual(values = c("blue", "red"), name = "D") +
  theme_minimal() +
  labs(x = "x (centered)", y = "y", title = "brms RD model fit")


```

# Multiple group regression discontinuity

work in progress

```{r}

```

# Session info

```{r sessioninfo}
sessionInfo()
```
