---
title: "Simulating error in Frequentist and Bayesian clinical trial designs"
date: last-modified

abstract: >
  Simulations of error under Frequentist and Bayesian frameworks. Technically, in Bayesian framework, there is no type I error, instead, the error is the probability of making a mistake. 
     
format: 
  html:
    toc: TRUE
    toc-depth: 4
editor: visual
editor_options: 
  chunk_output_type: console
---

# Load libraries

```{r load libraries, warning=FALSE,message=FALSE}
library(tidyverse)
library(brms)
library(gt)
library(gtsummary)
brms_summary <- function(x) {
  posterior::summarise_draws(x, "mean", "sd",  ~quantile(.x, probs = c(.025,0.975)))
}
```

# Introduction

Type I error in a Frequentist framework is rejecting the null hypothesis when the null hypothesis is true. Type I error does not apply in Bayesian framework, and instead, the focus on making an error when acting. This is making making a mistake in decision making, and can be thought of as 1-pr(benefit) (as an example).

I'll apply simulations to assessing error rate under frequentistism (type I error) as well as under a Bayesian framework (probability of making a mistake). I want to compare 2 priors: one with large standard deviation ($Normal(0,100)$) vs one with a narrower standard deviation ($Normal(0,5)$).

Probability of benefit is defined as $Pr(\beta)>0.95$

Frequentist approach steps:

1.  Simulate 1000 datasets with a mean difference of 0, sd of 1 with difference sample sizes (50,100,200) 2.Approach for Frequentist type I error: For each dataset, Fit Bayesian model with two priors (N(0,5) and N(0,100))
2.  Obtain 95% credible interval and median 
3.  Calculate how many times Pr(Benefit\>0.95)

Bayesian approach steps:

1.  Simulate 1000 datasets with a mean difference of 4, sd of 6 with difference sample sizes (50,100,200)
2.  Approach for Bayesian type I error: For each dataset, fit bayesian model with two priors (N(0,5) and N(0,100))
3.  Obtain 95% credible interval and median
4.  Calculate how many times 1-Pr(Benefit\>0.95) or Pr(harm)\>0.05

Then, compare error for each approach by sample size and prior.

# Frequentist simulation code

```{r frequentist, eval=FALSE}

#fit initial model to update
n=100
df<-tibble(treatment=rbinom(n,1,.5))|>
    rowwise()|>
    mutate(y=ifelse(treatment ==0,rnorm(n=1,0,6),
                            rnorm(n=1,0,6)))

ffitf<-brm(y~treatment,data=df,prior=prior(normal(0,5)))# prior with N(0,5)
ffit1000f<-brm(y~treatment,data=df,prior=prior(normal(0,1000)))# prior with N(0,1000)


#simulation function  for N(0,5 ) and N(0,1000) prior
simfit3 <- function(seed, n) {
  
  set.seed(seed)
  
  df<-tibble(treatment=rbinom(n,1,.5))|>
    rowwise()|>
    mutate(y=ifelse(treatment ==0,rnorm(n=1,0,6),
                            rnorm(n=1,0,6)))
  outp<-update(ffitf,
         newdata = df,
         seed = seed,cores=4,recompile=FALSE,refresh=0,prior=prior(normal(0,5))) 
  ot<-outp|>
    brms_summary()|>
  data.frame()|>
  filter(variable=="b_treatment")
  benefit<-hypothesis(outp,"treatment >0",alpha=0.05)
  ot$benefit<-benefit$hypothesis$Post.Prob
  ot$prior<-5
  ###
  outp2<-update(ffit1000f,
         newdata = df,
         seed = seed,cores=4,recompile=FALSE,refresh=0,prior=prior(normal(0,1000))) 
  ot2<-outp2|>
    brms_summary()|>
    data.frame()|>
    filter(variable=="b_treatment")
  benefit2<-hypothesis(outp2,"treatment >0",alpha=0.05)
  ot2$benefit<-benefit2$hypothesis$Post.Prob
  ot2$prior<-100
  
  ot3<-rbind(ot,ot2)
  
  return(ot3)
}
```

## Frequentist type I error Simulation

Many don't agree with this, but we're simulating outcomes under the null hypothesis with Bayesian models. Then, we can identify how often positive results are observed over the simulation.

```{r}
###--Actual simulation--###
#not run because it takes a very long time
##ss is set to different sample sizes 
#n_sim<-100 # number of simulations
#sim3<-tibble(expand.grid(seed=1:n_sim,ss=c(25,50,100,200,500)))|>
#  group_by(seed,ss)|>
#  do(out=simfit3(seed=.$seed,n=.$ss))|>
#  unnest(out)

#head(sim3)
#saveRDS(sim3,"Power_analysis_simulations_outputs/Type_1_error_Frequentist_framework_priorN05_N01000.rds")
sim3<-readRDS("Power_analysis_simulations_outputs/Type_1_error_Frequentist_framework_priorN05_N01000.rds")

sim3.parse<-sim3|>
  group_by(seed,ss,prior)|>
  mutate(benefit_yesno=if_else(X2.5.>0,1,0))|>
  group_by(ss,prior)|>
  dplyr::summarize(mean=mean(benefit_yesno))

sim3.parse

ggplot(sim3.parse,aes(x=ss,y=mean,colour=factor(prior)))+geom_point(size=5)+geom_line()+scale_y_continuous(limits=c(0,.06),breaks=seq(0,0.06,.01),labels=seq(0,.06,.01))+geom_hline(lty="dotdash",yintercept = 0.025)+ylab("Type I Error")+xlab("Sample size")+scale_x_continuous(limits=c(0,500),breaks=seq(0,500,50),labels=seq(0,500,50))

```

# Bayesian

## The simulation function

```{r bayesian, eval=FALSE}

#fit initial model to update
n=100
d<-tibble(treatment=rbinom(n,1,.5))|>
    rowwise()|>
    mutate(y=ifelse(treatment ==0,rnorm(n=1,0,6),
                            rnorm(n=1,4,6)))

ffit<-brm(y~treatment,data=d,prior=prior(normal(0,5)))# prior with N(0,5)
ffit1000<-brm(y~treatment,data=d,prior=prior(normal(0,1000)))# prior with N(0,1000)
#hypothesis(ffit,"treatment>0",alpha=0.05)



#simulation function  for N(0,5 ) and N(0,1000) prior
simfit <- function(seed, n) {
  
  set.seed(seed)
  
  d<-tibble(treatment=rbinom(n,1,.5))|>
    rowwise()|>
    mutate(y=ifelse(treatment ==0,rnorm(n=1,0,6),
                            rnorm(n=1,4,6)))
  outp<-update(ffit,
         newdata = d,
         seed = seed,cores=4,recompile=FALSE,refresh=0,prior=prior(normal(0,5))) 
  ot<-outp|>
    brms_summary()|>
  data.frame()|>
  filter(variable=="b_treatment")
  benefit<-hypothesis(outp,"treatment >0",alpha=0.05)
  ot$benefit<-benefit$hypothesis$Post.Prob
  ot$prior<-5
  ###
  outp2<-update(ffit1000,
         newdata = d,
         seed = seed,cores=4,recompile=FALSE,refresh=0,prior=prior(normal(0,1000))) 
  ot2<-outp2|>
    brms_summary()|>
    data.frame()|>
    filter(variable=="b_treatment")
  benefit2<-hypothesis(outp2,"treatment >0",alpha=0.05)
  ot2$benefit<-benefit2$hypothesis$Post.Prob
  ot2$prior<-100
  
  ot3<-rbind(ot,ot2)
  
  return(ot3)
}

```

## Bayesian simulation

```{r actual bayesian simulation}
#n_sim<-100 # number of simulations
###--##comenting out simulation code - takes very long time
#sim2<-tibble(expand.grid(seed=1:n_sim,ss=c(25,50,100,200,500)))|>
#  group_by(seed,ss)|>
#  do(out=simfit(seed=.$seed,n=.$ss))|>
#  unnest(out)


#saveRDS(sim2,"Power_analysis_simulations_outputs/Type_1_error_Bayesian_framework_priorN05_N01000.rds")
sim2<-readRDS("Power_analysis_simulations_outputs/Type_1_error_Bayesian_framework_priorN05_N01000.rds")|>
  mutate(prior=if_else(prior==100,1000,prior))
head(sim2)

```

### Side quest- Generate a function that can determine power for a given sample size


Grab data and run on your own.  

```{r bayesian power, eval=FALSE}
#Get power 
sim2.parse<-sim2|>
  group_by(seed,ss)|>
  mutate(benefit_yesno=if_else(X2.5.>0,1,0))|>
  group_by(ss,prior)|>
  dplyr::summarize(mean=mean(benefit_yesno))

sim2.parse

##plot out the power by 
ggplot(sim2.parse,aes(x=ss,y=mean,colour=factor(prior)))+geom_line()+geom_point()+scale_x_continuous(limits=c(0,500),breaks=seq(0,500,50),labels=seq(0,500,50))


########---### create a power function so we can predict power given a sample size input
sim2.parse5<-sim2.parse|>
  filter(prior==5)
# Hill model: y = ss^n / (K^n + ss^n)
# Use non-linear brms formula syntax

hill_formula <- bf(
  mean ~ ss^n / (K^n + ss^n),
  n + K ~ 1,
  nl = TRUE
)

priors <- c(
  prior(normal(2, 1), nlpar = "n", lb = 0),
  prior(normal(60, 30), nlpar = "K", lb = 0),
  # Use a proper prior for sigma on (0, ∞). Pick one:
  prior(student_t(3, 0, 0.2), class = "sigma", lb = 0)   # robust, fairly tight
)

fit_bayes <- brm(
  formula = hill_formula,
  data = sim2.parse5,
  family = gaussian(),       # still Gaussian
  prior = priors,
  control = list(adapt_delta = 0.95),
  chains = 4, cores = 4
)

summary(fit_bayes)

# Create a tidy posterior prediction function
predict_hill <- function(ss_values) {
  newdata <- data.frame(ss = ss_values)
  posterior_epred(fit_bayes, newdata = newdata)
}

# Example: predict for ss = 75
mean(predict_hill(75))


##predict continuously from 0 to 500 

powerf<-apply(predict_hill(seq(0,500,1)),2,mean)
q025<-apply(predict_hill(seq(0,500,1)),2,quantile,probs=0.025)
q975<-apply(predict_hill(seq(0,500,1)),2,quantile,probs=0.975)
powerf1<-tibble(mean=powerf,ss=seq(0,500,1),lci=q025,uci=q975)
###--plot with simulation points 

ggplot(sim2.parse5,aes(x=ss,y=mean))+geom_point()+scale_x_continuous(limits=c(0,500),breaks=seq(0,500,50),labels=seq(0,500,50))+geom_line(data=powerf1,aes(x=ss,y=mean))+geom_ribbon(data=powerf1,aes(x=ss,ymin=lci,ymax=uci),alpha=.2)+xlab("Sample size")+ylab("Bayesian power Pr(benefit)>=0.95")

powerf1|>
  filter(ss==77)

```

### Back to original goal: figure out pr(harm)

Calculate the pr(harm), which is 1-pr(benefit) for our simulations

```{r}
harmd<-sim2|>
  mutate(harm=1-benefit)|>
  group_by(ss,prior)|>
  summarize(mean=mean(harm),lcri=quantile(harm,prob=0.025),ucri=quantile(harm,0.975))
harmd

#plot out 
ggplot(harmd,aes(x=ss,y=mean,colour=factor(prior),fill=factor(prior)))+geom_point()+geom_line()+geom_ribbon(alpha=0.2,aes(ymin=lcri,ymax=ucri),colour=NA)+scale_x_continuous(labels=seq(0,500,50),breaks=seq(0,500,50),limits=c(0,500))+scale_y_continuous(breaks=seq(0,.7,.05),labels=seq(0,.7,.05))+geom_hline(yintercept = 0.025,lty="dotdash")+xlab("Sample size")+ylab("Pr(harm) or Pr(treatment<0)")


```

So a prior with narrower sd produces a higher error rate than a larger sd prior.

# Session info

```{r sessioninfo}
sessionInfo()
```
