---
title: "Ordinal regression modeling"
format: 
  html:
      toc: true
      toc-depth: 3
editor: visual
editor_options: 
  chunk_output_type: console
---

# Load libraries

```{r loading libraries, message=FALSE,warning=FALSE}
library(tidyverse) #data parsing and visualization 
library(gtsummary) #this lets you create nice tables
library(gt) # create nice tables
library(brms) # bayesian modeling
#library(REDCapDM) # interacting with redcap API
library(splines)
library(marginaleffects)
#library(survival)
library(ggbeeswarm)
library(patchwork)
library(tidybayes)


brms_summary <- function(x) {
  posterior::summarise_draws(x, "mean", "sd",  ~quantile(.x, probs = c(.025,0.975)))
}
```

# Introduction

In this script, I'm going to simulate different ordinal outcomes to explore how the ordinal regression modelling works. Some ideas to explore-

-   Simply analyzing a continuous outcome as ordinal - Frank Harrell says there is a lot of power by doing this.\
-   5 ordinal scale - 1-4 is qol and 5 is death
    -   Get outcomes in terms of probability 1, \<2, \<4, 5
-   Simulate proportional odds and non-proportional odds: evaluate models
-   Explore longitudinal aspect

## Notes on ordinal regression models

Taking notes from ordinal regression in brms [github](https://github.com/kalealex/ordinal-regression-demo).

Ordinal data are categories that show grades of an outcome. Ordinal regression models (cumulative models) assumes that the ordinal variable comes from a categorization of a latent (not observed) continuous $\hat{Y}$ variable.

Example of ordinal data:

```{r ordinal dataset}
knitr::kable(tibble(groups=c("fundamentalist","moderate","liberal"),`1`=c(40,25,23),`2`=c(54,41,31),`3`=c(119,135,113),`4`=c(55,71,122)))
```

Assuming 4 categories:

There are $K$ thresholds $\tau_k$ which partition $\hat{Y}$ into K+1 observable, ordered categories of $Y$. With 4 categories assumed, there are K+1=4 response categories, and K=3 thresholds.

Assuming $\hat{Y}$ follows a normal distribution with a cumulative distribution function $F$ (CDF), then the probability of $Y$ being equal to category $k$ is:

$$Pr(Y=k)= F(\tau_k)-F(\tau_{k-1})$$ The CDF of a threshold is subtracted from the CDF of the next lower level threshold to obtain the probability of Y being in a certain category.

Example: $\phi$ is a normal CDF

$$Pr(Y=2)=\phi(\tau_2) - \phi(\tau_1) = \phi(1)-\phi(-1)= 0.84 -0.16 = 0.68$$

We need to incorporate and describe a regression model. So, the linear regression for $\hat{Y}$ with predictors is $\eta=b_1x_1+b_2x_2+..,$, so that $\hat{Y}= \eta + \epsilon$, where $\epsilon$ is the error term in the regression. Then, $\hat{Y}$ can be split into these two parts. Note that there is no intercept in the predictor term because the thresholds $\tau_k$ replace the model's intercept as both are not identified at the same time. Thus, the probabilities of $Y$ being equal to category $k$ given the linear predictor $\eta$ is:

$$Pr(Y=k|\eta)= F(\tau_k-\eta) - F(\tau_{k-1}-\eta)$$

Thus..

$$\hat{Y} = \eta + \epsilon = b_1 x_1 + b_2 x_2 + \epsilon$$ Assuming the latent variable $\hat{Y}$ is normally distributed, then the probabilities for each response cateogry $k$ can be computed as:

$$Pr(Y=k) = \phi(\tau_k-(b_1x_1 + b_2x_2))-\phi(\tau_{k-1}-(b_1x_1 + b_2 x_2))$$

There are also other models: sequential model (where movement across ordinal values depends on the last ordinal value), and the adjacent category model. Appendix A has more details.

### Practical application notes

-   We can specify a few different models, all of which can be compared with leave-one-out cross validation (LOO).
    -   If comparing ACM, CM, and SM models, especially with category specific effects ($cs()$), then LOO should be also compared with model not specified with $cs()$.\
-   brms can explore and account for unequal variances that can vary with predictors (but not cut off values).\
-   brms can specify different levels of effects across different response categories. This is specified by $cs()$ function. Category specific estimates are not problematic for SM and ACM models, but there could be negative probabilities for cumulative models.
-   Rhat and ESS are checked for chain convergence. Rhat should not be above 1.1 and ESS should be as high as possible. ESS \>1000 is a rule of thumb for stable estimates.

# Let's first simulate a dataset for proportional ordinal regression

```{r simulation proportional}
#Let's set up probabilities, then add in some simulate outcomes based on those probabilities 
ss<-1000 # sample size
control.prob<-c(0.1,.2,.3,.25,.15)
#sum(control.prob)# has to add up to 1
OR<-.5 # odds ratio

#creat cdf for treatment which is a function of control cdf that has an OR applied to it
t.prob<-plogis(qlogis(cumsum(control.prob)[1:4]) - log(OR))
#from cdf, back calculate to probabilities
t.prob1<-tibble(prob=c(t.prob[1],diff(t.prob),1-t.prob[4]))

#sample data ; control and treatment
#control 
control<-sample(size=ss,x=1:5,prob=control.prob,replace=TRUE)
#treatment 
trt<-sample(size=ss,x=1:5,prob=t.prob1$prob,replace=TRUE)

#merge dataset 
d<-tibble(y=c(control,trt),treat=c(rep("control",ss),rep("treated",ss)))
head(d)
#levels(factor(d$treat))
#let's plot the data 

d|>
  count(treat,y)|>
  mutate(prob=n/ss)|>
  ggplot(aes(x=y,y=prob,colour=treat))+geom_point(size=4)+geom_line()+theme_bw()+theme(legend.position="top")
```

## Now let's fit a proportion ordinal regression model

Let's use skeptical prior - normal(0,log(5)/1.96)

```{r po ordinal regression model }
# set Priors:
priors <- c(
  set_prior("student_t(3, 0, 5)", class = "Intercept"),  # cutpoints θ1..θ(K-1)
  set_prior("normal(0, 0.8211418)",       class = "b")           # slopes on log-odds
)

#pomod1 <- brm(
#  y ~ treat,                     # add predictors as needed
#  data   = d,
##  family = cumulative("logit"),     # proportional odds
#  prior  = priors,
#  chains = 4, iter = 4000, seed = 123
#)
#let's save the model so we don't have to rerun 
#saveRDS(pomod1,file="31_proportional_ord_reg_model_Bayesian_prior0.2-5.0_COR_pomod1.rds")
#read in the model 
pomod1<-readRDS(file="31_proportional_ord_reg_model_Bayesian_prior0.2-5.0_COR_pomod1.rds")
summary(pomod1) # get Rhat, ESS
#plot(pomod1) # trace and densities
pp_check(pomod1,type="bars",ndraws=100) # posterior predictive vs observed

# get model summary - mean and 95% CrI
brms_summary(pomod1)|>
  filter(variable=="b_treattreated")|>
  select(-variable)|>
  exp()
#very close to the real estimate, COR = 0.5!
#if close to 1, can test hypotheses about being above or below 1 COR


#chekc proportional odds 
# approach - fit with a partial proportional odds model 
# here, we suspect that the treatment effect differs by threshold 
#then we use leave one out to determine which model is better 

#pomod.partial <- brm(
#  y ~ cs(treat),                     # add predictors as needed
#  data   = d,
#  family = cumulative("logit"),     # proportional odds
#  prior  = priors,
#  chains = 4, iter = 4000, seed = 123
#)
#save and read in pomod.partial
#saveRDS(pomod.partial,file="31_partial_proportional_ord_reg_model_Bayesian_prior0.2-5.0_COR_pomod.partial.rds")
pomod.partial<-readRDS(file="31_partial_proportional_ord_reg_model_Bayesian_prior0.2-5.0_COR_pomod.partial.rds")

summary(pomod.partial)
#leave one out analysis 
#it is the out-of-sample predictive accuracy 
#It asks, which model is better at predicting unseen data?

##output
#elpd_loo → Expected log predictive density
#     * higher means better predictive perofrmance , the difference between models matter 
#se_elpd_loo → Standard error of elpd
#looic → LOO information criterion (−2 * elpd_loo)
#p_loo → Effective number of parameters
#pareto_k → Diagnostics


#loopo<-loo_compare(loo(pomod1),loo(pomod.partial))
#saveRDS(loopo,file="31_LOO_CMmodels_proportional_2models.rds")
loopo<-readRDS(file="31_LOO_CMmodels_proportional_2models.rds")
knitr::kable(loopo)
```

pomod1 predicts new data better than pomod.partial (higher ELPD is better; the top row is set to 0 by definition and is the best).

The difference in ELPD is 2.5 with SE = 1.0. A common rule-of-thumb:

if \|elpd_diff\| \> 2 × se_diff, there is strong evidence the top model predicts better.

Here, 2.5/1.0=2.52.5 / 1.0 = 2.52.5/1.0=2.5 ⇒ strong evidence in favor of pomod1.

## Extract predicted probabilities for each category

### Not very tidy way of extraction

```{r non tidy way of data parsing,eval=FALSE}
# predict cateogry probabilities
p_cat <- posterior_epred(pomod1, newdata = tibble(treat=c("control","treated")))  # draws x rows x K
# Posterior means per arm and category:
apply(p_cat, c(2,3), mean)          # rows: newdat rows, cols: categories
# 95% CrI:
apply(p_cat, c(2,3), quantile, c(.025,.975))
#not very tidy...
```

### Using marginaleffects to extract

This is much cleaner

```{r using marginaleffects }
#set up new data
newd<-tibble(treat=c("control","treated"))
#get marginal probabilities by treatment 
avg_pr<-avg_predictions(pomod1,variables="treat",type="response",newdata=newd)
head(avg_pr)
#let's plot 

fig1<-ggplot(avg_pr,aes(y=estimate,x=group,colour=treat,group=treat,fill=treat))+geom_line()+geom_point(size=3)+geom_ribbon(aes(ymin = conf.low,ymax=conf.high),alpha=.1,colour=NA)+theme_bw()+xlab("ordinal scale")+ylab("Probability")+theme(legend.position = "top")
#looks like the original dataset, which is good 

#lets get average differences 
avg_diff<-avg_comparisons(pomod1,variables="treat")
head(avg_diff)
fig2<-avg_diff|>
  data.frame()|>
  select(-c(term,contrast))|>
  ggplot(aes(x=group,y=estimate,group=1))+geom_ribbon(aes(ymin = conf.low,ymax=conf.high),alpha=.1,colour=NA)+geom_point()+geom_line()+theme_bw()+geom_hline(yintercept = 0,lty="dotdash")+xlab("ordinal scale")+ylab("ATE (Treatment-Control)")

(fig1|fig2)+plot_annotation(tag_levels = "A")

```

# Partial proportional ordinal regression model- simulate data first

Let's say the control mostly die (category 5) compared to treated. So I'll set the control and treatment to have their own probabilities

```{r simulation partial }
#Let's set up probabilities, then add in some simulate outcomes based on those probabilities 
ss<-1000 # sample size
control.prob<-c(0.1,.2,.3,.25,.15)
#sum(control.prob)# has to add up to 1
treat.prob<-c(0.025,.025,.025,.025,.9)
#sample data ; control and treatment
#control 
control<-sample(size=ss,x=1:5,prob=control.prob,replace=TRUE)
#treatment 
trt<-sample(size=ss,x=1:5,prob=treat.prob,replace=TRUE)

#merge dataset 
d2<-tibble(y=c(control,trt),treat=c(rep("control",ss),rep("treated",ss)))
head(d2)
#levels(factor(d$treat))
#let's plot the data 

d2|>
  count(treat,y)|>
  mutate(prob=n/ss)|>
  ggplot(aes(x=y,y=prob,colour=treat))+geom_point(size=4)+geom_line()+theme_bw()+theme(legend.position="top")
```

## Fit a partial proportional ordinal regression model

```{r fit partial model}
# set Priors:
priors2 <- c(
  set_prior("student_t(3, 0, 5)", class = "Intercept"),  # cutpoints θ1..θ(K-1)
  set_prior("normal(0, 0.8211418)",       class = "b")           # slopes on log-odds
)

######part.mod1 fit 
#part.mod1 <- brm(
#  y ~ cs(treat),                
#  data   = d2,
#  family = cumulative("logit"),     
#  prior  = priors2,
#  chains = 4, iter = 4000, seed = 123
#)
#saveRDS(part.mod1,file="31_part.mod1_bayesian_partialpomod_case_specificfit.rds")
part.mod1<-readRDS(file="31_part.mod1_bayesian_partialpomod_case_specificfit.rds")
summary(part.mod1)

pp_check(part.mod1,type="bars",ndraws=100) # posterior predictive vs observed

# get model summary - mean and 95% CrI
brms_summary(part.mod1)|>
  filter(grepl(variable,pattern="bcs_")==TRUE)|>
  mutate(mean=exp(mean),`2.5%`=exp(`2.5%`),`97.5%`=exp(`97.5%`))|>
  select(-sd)

conditional_effects(part.mod1,categorical=TRUE)

########ACM ordinal regression fit #####################

#the tutorial for ordinal regression in brms recommends different specification: 

#case specific effects
#part.mod2 <- brm(
#  y ~ cs(treat),                
#  data   = d2,
#  family = acat("probit"),     
#  prior  = priors2,
#  chains = 4, iter = 4000, seed = 123
#)
#saveRDS(part.mod2,file="31_part.mod2_bayesian_ACMmod_probit_specificfit.rds")
part.mod2<-readRDS("31_part.mod2_bayesian_ACMmod_probit_specificfit.rds")
summary(part.mod2)
conditional_effects(part.mod2,categorical=TRUE)

#proportional effects 
#part.mod3 <- brm(
#  y ~ treat,                
#  data   = d2,
#  family = acat("probit"),     
#  prior  = priors2,
#  chains = 4, iter = 4000, seed = 123
#)
#saveRDS(part.mod3,file="31_part.mod3_bayesian_ACMmod_probit_NO_specificfit.rds")
part.mod3<-readRDS("31_part.mod3_bayesian_ACMmod_probit_NO_specificfit.rds")
summary(part.mod3)
conditional_effects(part.mod3,categorical=TRUE)
#######

###now comparing with proportional ordinal regression model 
#Leave one out cross validation 
#In the context of model selection, a LOO difference greater than twice its corresponding
#standard error can be interpreted as suggesting that the model with a lower LOO value fits
#the data substantially better, at least when the number of observations is large enough
#loopartial<-loo_compare(loo(part.mod1),loo(part.mod2),loo(part.mod3))
#saveRDS(loopartial,file="31_LOO_partial_proportional_3models.rds")
loopartial<-readRDS(file="31_LOO_partial_proportional_3models.rds")
knitr::kable(loopartial)
```

The case specific effects models are more favorable models. Out of the case specific effects models, specifying with the model is ACM (acat) or CM yield similar models.

# Now, analyzing a continuous variable with ordinal regression

```{r simulate continuous data and convert to ordinal scale}

co<-rnorm(n=ss,50,5)
t<-co+rnorm(n=ss,5,1)
cond<-tibble(y=c(co,t),treat=c(rep("control",ss),rep("treated",ss)))
ggplot(cond,aes(y=treat,x=y,fill=treat))+stat_halfeye() 
#let's convert to ordinal scale 
cond$yord<-factor(round(cond$y,0),ordered=TRUE)

##set up prior 
priors3 <- c(
  set_prior("student_t(3, 0, 5)", class = "Intercept"),  # cutpoints θ1..θ(K-1)
  set_prior("normal(0, 2.349577)",       class = "b")           # slopes on log-odds
)
# I don't know what is plausible for cumulative odds ratio -> so I'll use a diffuse prior log(100)/1.96

#let's fit proportional CM model 
#comod1 <- brm(
#  yord ~ treat,                    
#  data   = cond,
#  family = cumulative("logit"),     # proportional odds
# prior  = priors3,
#  chains = 4, iter = 4000, seed = 123,control = list(adapt_delta = 0.98)
#)
#saveRDS(comod1,file="31_comod1_ord_reg_CMmod_continuous_outcome_converted_to_ordinal.rds")
comod1<-readRDS(file="31_comod1_ord_reg_CMmod_continuous_outcome_converted_to_ordinal.rds")
summary(comod1) # get Rhat, ESS
#plot(comod1)
pp_check(comod1,type="bars",ndraws=100) # posterior predictive vs observed

# get model summary - mean and 95% CrI
brms_summary(comod1)|>
  filter(variable=="b_treattreated")|>
  select(-variable)|>
  exp()
# The cumulative odds is 5.58 and this means that the treated increases the odds by 5.58 to being in a higher ranked category. 
conditional_effects(comod1,categorical=TRUE)
#how many levels do i ahve? 
length(levels(comod1$data$yord))

```

Notes- ordinal regression can work on continuous data, which means that you can break a continuous scale into ordinal values. This may be important for cases where you want to merge a continuous outcome with a terminal, absorbing state (death). In our case, 71 was highest continuous scale, and if we wanted to include death, then 72 could be representing death.

To better obtain convergence of chains, the adapt_delta was set to 0.98. At adapt_delta set to 0.95, there were divergent chains.

# Session info

```{r session information}
sessionInfo()
```
