---
title: "Publication Quality Figures -> ggplot2"
author: "Andrew Nguyen"
date: "2023-08-20"
output: 
  html_document: 
    theme: cosmo
    toc: yes
editor_options: 
  chunk_output_type: console
---

I'll be demo'ing the tidyverse on a project I worked on in my postdoc at the Unversity of Florida. The project details are publicly available through this github repository. 


# Load libraries 

```{r, load libaries}
library(tidyverse) # for ggplot2, data visualization and data filtering with dplyr 
library(ggbeeswarm) # for quasirandom plotting
library(caret)# for ML analysis

#ggplot2 settings I like: 
T<-theme_bw()+theme(,text=element_text(size=25),axis.text=element_text(size=18), panel.grid.major=element_blank(), panel.grid.minor.x = element_blank(), panel.grid = element_blank(), legend.key = element_blank(),axis.title.y=element_text(margin=margin(t=0,r=15,b=0,l=0)),axis.title.x=element_text(margin=margin(t=15,r=,b=0,l=0)))#+ theme(legend.position="none")
```

## The dataset! Let's explore different plots

There's a nice dataset I generated from my postdoctoral research in Dr. Dan Hahn's lab. I collected eclosion times (when an insect transitions from pupae to adult) for fruit flies (Rhagoletis) from either the apple or hawthorne fruit. The project and details are found [here](https://github.com/adnguyen/Circadian_rhythm_runs_seasonal_timing) and has metadata. 

```{r fruit fly}
#fruit.fly<-read.delim("https://raw.githubusercontent.com/adnguyen/Circadian_rhythm_runs_seasonal_timing/master/Data/2018-08-14_master_data_slice.csv")

fruit.fly<-read.csv("https://raw.githubusercontent.com/adnguyen/Circadian_rhythm_runs_seasonal_timing/master/Scripts_analyses/05_master_dataset_eclosions_IDs.csv")
#glimpse(fruit.fly)
#take out the data of interest
fruit.fly1<-fruit.fly%>%
  select(Site_name,mass_day10,Host,cohort_day,resp_day15,new.eclosions,organism,treatment)%>%
  dplyr::filter(organism=="fly" &treatment !="GC" & treatment!="")# GC = genetic controls, those were saved for genomic analyses and we're just focusing on eclosion of flies (not the parasites)
glimpse(fruit.fly1)
```





I need to find out proportions of eclosion and need to transform eclosion data such that NA = 0 and a number in days = 1. And I want to display the proportions by treatment (RT = Room Temperature vs SO = Simulated Overwintering) and host fruit (apple vs haw). 

```{r data manipulation}


fruit.fly1$eclfac<-as.numeric(ifelse(fruit.fly1$new.eclosions>1,"1","0"))
fruit.fly1$eclfac[is.na(fruit.fly1$eclfac)]<-0

ecl.num<-fruit.fly1%>%
  group_by(treatment,Host,eclfac)%>%
  count()
knitr::kable(ecl.num)

```

**You can see most flies have eclosed.**



We can show the data in a number of different ways:     

* barplots    
* cumulative distribution plots   


### Barplots   

Let's first start off with a barplot:

```{r barplot}

#let's create a barplot 
# treatments, RT = room temperature, SO = simulated overwintering 
ggplot(ecl.num,aes(y=n,x=factor(eclfac),fill=factor(eclfac)))+
  geom_bar(stat="identity")+
  facet_wrap(treatment~Host)+ #this function creates multi-panels and is super handy at conveying complex datasets with multiple treatments /categorical variables
  ylab("Number of Adults Emerged")+xlab("Eclosed (1) or not (0)")+
  T+ # my own plotting settings, see library
  theme(legend.position="none")+ # removing legend
  scale_y_continuous(expand=c(0,0),limits=c(0,300),breaks=seq(0,300,100),labels=seq(0,300,100))+ # makes sure the y-axis starts at 0, this is something I always use when making barplots with ggplot2
  geom_text(aes(label=n),size=5,vjust=-.5) # it is nice to annotate the bars to help the eyes interpret them
```

Let's remove the ones that have not eclosed and observe their differences in timing and we can visualize the data as a cumulative distribution plot. 
==Remember to save the figure with ggsave() and you can specify the dpi and dimensions of the plot.== 


```{r removing zeros and ECDF plots}
ff1<-fruit.fly1%>%
  dplyr::filter(eclfac==1)

ggplot(ff1,aes(x=new.eclosions,colour=Host))+stat_ecdf(linewidth=1)+facet_wrap(~treatment)+T+scale_x_continuous(limits=c(0,300),labels=seq(0,300,100),breaks=seq(0,300,100))+xlab("Days Until Adult Emergence (Days)")+ylab("ECDF")

```


### Quasirandom plots 

These types of plots are nice to show the distribution of the data. In this case, we have continuous data on the y axis and categorical on the x. 
```{r quasi random plots}

ggplot(ff1,aes(x=Host,y=new.eclosions,fill=Host))+
  geom_quasirandom(size=2,alpha=.5,shape=21,colour="black")+ # change the points so that there is a black outline and color is filled in by host
  facet_wrap(~treatment)+ylab("Days Until Adult Emergence (Days)")+
  stat_summary(fun = mean, geom = "errorbar", aes(ymax =after_stat(y) , ymin = after_stat(y)),width = .75,lwd=1,colour="black")#add average horizontal line, which adds a nice touch

```

It looks like there are two populations within the RT treatment. When conditions are optimal for growth (they're at room temperature and not diapausing), then there looks to be a subpopulation that emerges very quickly. 


### Scatter plots to display regressions

For illustrative purposes, I'm plotting mass with eclosion days just to show how to implement scatter plots. 

```{r scatter plots}
ggplot(ff1,aes(x=mass_day10,y=new.eclosions,colour=Host))+geom_point()+stat_smooth(method="lm")+#fits regression lines with standard errors and takes into account the Host when colour is specified
  facet_wrap(~treatment,scale="free")+ylab("Days Until Adult Emergence (days)")+xlab("Mass (mgs)")

```

## Survival analysis

The data are best suited for a surival analysis because we have time of events. 

```{r survival analysis}

```

## Machine Learning -> predicting host fruit from organismal features

Is it possible to predict host fruit based on the measurements we made? We'll create data partitions for training and then testing the model. 

```{r SVM }
# I need to convert Host into a factor
ff1$hfac<-factor(ifelse(ff1$Host=="Apple","1","0"))

ff2<-ff1%>%
  dplyr::filter(treatment=="SO")%>%
  dplyr::select(hfac,mass_day10,resp_day15,new.eclosions)
  #just picking out variables that we want 

intrain <- createDataPartition(y = ff2$hfac, p= 0.7, list = FALSE)
training <- ff2[intrain,]
testing <- ff2[-intrain,]

trctrl <- trainControl(method = "repeatedcv", number = 10, repeats = 3)
#The “method” parameter defines the resampling method, in this demo we’ll be using the repeatedcv or the repeated cross-validation method.

#The next parameter is the “number”, this basically holds the number of resampling iterations.

#The “repeats ” parameter contains the sets to compute for our repeated cross-validation. We are using setting number =10 and repeats =

svm_Linear <- train(hfac ~., data = training, method = "svmLinear",
trControl=trctrl, preProcess = c("center", "scale"), tuneLength = 10)

svm_Linear
##let's see how well it predicts

test_pred <- predict(svm_Linear, newdata = testing)
confusionMatrix(table(test_pred, testing$hfac))
```


# Session info

```{r, sessioninfo}
sessionInfo()
```