---
title: "Title - SQL demo"
author: "Author: Andrew D. Nguyen, PhD, Quantitative Biologist"
date: "Date:2023-08-27"
output: 
  pdf_document: 
    toc: yes
    toc_depth: 5
    number_sections: yes
editor_options: 
  chunk_output_type: console
---

\newpage

# Introduction -> Goals   

In this demo, I become more familiar with interacting with SQL databases. I'll be following a tutorial from **[data carpentry](https://datacarpentry.org/R-ecology-lesson/05-r-and-databases.html)**, but I'd also like to peruse open databases to become more versed in SQL and querying databases into R for data wrangling and analyses. Lastly, I'll construct my own databases from datasets I have generated.    

# Load Libraries

```{r load libaries, message=FALSE}
library(tidyverse) # data wrangling
library(RSQLite) # SQL - R package
library(dbplyr) # SQL - R package, lets dplyr apply to
#SQL databases
```

# The database - Portal Mammals Database

I have to first download the portal mammals database off of figshare. Then, load the mammals database.

```{r portal mammals database}
#download.file(url = "https://ndownloader.figshare.com/files/2292171",
#              destfile = "data/portal_mammals.sqlite", mode = "wb")
# load data
mammals<-DBI::dbConnect(RSQLite::SQLite(), "data/portal_mammals.sqlite")
#This command does not load the data into the R session
#(as the read_csv() function did). 
#Instead, it merely instructs R 
#to connect to the SQLite database
#contained in the portal_mammals.sqlite file.

#now, lets see what the database is made of
src_dbi(mammals)
```



# SessionInfo

```{r session info}
sessionInfo()

```